Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-28 paper/paper.tex
Fixed as per discussion with Prof.AS

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}  and  Haskell~\cite{rojemo96lag}
programs have shown that garbage  collectors leave uncollected a large
number of  memory objects that are  reachable but not live  (here {\em
  live} means the  object can potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

@@

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) &  \mbox{\em --- selects $1^{st}$ part of \CONS} \\
       (\CDR\,\, x) & \mbox{\em --- selects $2^{nd}$ part of \CONS} \\
       (\NULLQ\,\, x) & \mbox{\em --- returns 0 if x is not $\NIL$}\\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
\vspace*{-6pt}
\end{figure}

@@

Here $\mathit{Loc}$  is a countable set  of locations in the  heap.  A
non-empty location either contains a {\em closure}, or a value in Weak
Head Normal Form (WHNF)\cite{Jones87}. In our case, a WHNF value
can be a number, the empty list  $\NIL$ or a \CONS\ cell with possibly
unevaluated   constituents.   A   closure  is   a  pair   $\langle  s,
\rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
maps free  variables of $s$  to their respective locations.  Since all
data objects are boxed, we model  an environment as a mapping from the
set of  variables of  the program $\mathit{Var}$  to locations  in the
heap.

@@

As  demonstrated by  the gap  between the  \RGCLine\ and  the \UseLine\
lines, a  large number of  cells which are  unused by the  program are
still copied  during RGC. LGC does  a much better job  of closing this
gap but still falls short of the  precision achieved by LGC in case of
eager languages ~\cite{asati14lgc}.  A major source of inefficiency in
LGC is  multiple traversals  of already copied  heap cells.  Since LGC
does not mark the heap cells after the first visit, the same cells can
be  repeatedly  visited  with   different  liveness  states.  We  have
mitigated  this  problem  by  adding checks to avoid multiple traversals 
of closure and function arguments.

>>>>

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}  and  Haskell~\cite{rojemo96lag}
programs have shown that garbage  collectors leave uncollected a large
number of  memory objects that are  reachable but not live  (here {\em
  live} means the  object may potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

@@

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) &  \mbox{\em --- selects first part of \CONS} \\
       (\CDR\,\, x) & \mbox{\em --- selects second part of \CONS} \\
       (\NULLQ\,\, x) & \mbox{\em --- returns true if x is $\NIL$}\\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
\vspace*{-6pt}
\end{figure}

@@

Here $\mathit{Loc}$  is a countable set  of locations in the  heap.  A
non-empty location either contains a {\em closure}, or a value in Weak
Head Normal Form (WHNF)\cite{Jones87}. For our implementation, a value
in WHNF is either a number, or  the empty list $\NIL$ or a \CONS\ cell
with possibly unevaluated constituents.  A  closure is a pair $\langle
s, \rho\rangle$ in which $s$ is an unevaluated application, and $\rho$
maps free variables  of $s$ to their respective  locations.  Since all
data objects are boxed, we model  an environment as a mapping from the
set of  variables of  the program $\mathit{Var}$  to locations  in the
heap.

@@

As   demonstrated  by   the  gap   between  the   \RGCLine\  and   the
\UseLine\  lines, a  large number  of cells  which are  unused by  the
program are  still copied during  RGC. LGC does  a much better  job of
closing this  gap but still falls  short of the precision  achieved by
LGC in case of eager  languages ~\cite{asati14lgc}.  A major source of
inefficiency  in LGC  is multiple  traversals of  already copied  heap
cells.  Since LGC does not mark  the heap cells after the first visit,
the  same cells  can  be repeatedly  visited  with different  liveness
states.  This can  be  seen in  the cells touched  per  GC for  $huffman$
benchmark in Table~\ref{fig:memory-usage}, even though it uses only 72
cells, number of cells touched is  88.  We have mitigated this problem
by adding checks to avoid  multiple traversals of closure and function
arguments.
